---
title: "SKILL_CHECK_3"
author: "OMARE_MOKAYA"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1&2 Loading Packages
```{r packages}

library(ggplot2)
library(patchwork)
library(dplyr)
library(tinytex)
```

## 3 Loading Data

```{r data}
gapminder <- read.csv("C:\\Users\\Omare Collins\\Downloads\\gapminder_data.csv")
glimpse(gapminder)
```

## 4 Scatter plot  of lifeExp vs gdpPercap

```{r scatter}
ggplot(gapminder,aes(x = gdpPercap, y = lifeExp))+
geom_point()
```

## 5 Colour by Continent
```{r colour}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point()
```

## 6 Trasforming the Y axis
```{r transform}
ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp, colour = continent)) +
geom_point()
```

## 7: Facet by continent

```{r facet}
ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp, colour = continent)) +
  geom_point() +
  facet_grid(~continent)
```

## 8: Theme as plot1

```{r plot1}
plot1 <- ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp, colour = continent)) +
  geom_point() +
  facet_grid(~continent) +
  theme_bw()
plot1
```

## 9: Density plot of life expectancy by continent

```{r density}
ggplot(gapminder, aes(x = lifeExp, fill = continent)) +
  geom_density(alpha = 0.5)
```

## 10: Life expectancy vs year, coloured by country, as plot2

```{r plot2}
plot2 <- ggplot(gapminder, aes(x = year, y = lifeExp, colour = country)) +
  geom_line() +
  theme(legend.position = "none")
plot2
```

## 11: gapminder_2007 - mean gdpPercap by continent

```{r}
gapminder_2007 <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarise(mean_gdpPercap = mean(gdpPercap))

gapminder_2007
# Oceania had the highest mean GDP per capita in 2007 (~$29,810),
# followed by Europe (~$25,054). Africa had the lowest (~$3,089).
```

## 12: Point plot of continent vs mean GDP per capita

```{r plot of ct vs gdp}
ggplot(gapminder_2007, aes(x = continent, y = mean_gdpPercap, colour = continent)) +
  geom_point()
```

## 13: Bar plot with labels and theme

```{r bar}
ggplot(gapminder_2007, aes(x = continent, y = mean_gdpPercap, colour = continent)) +
  geom_col() +
  labs(x = "Continent", y = "Mean GDP per Capita (2007)",
       title = "Mean GDP per Capita by Continent (2007)") +
  theme_minimal()
```

## 14: gapminder_summary - mean gdpPercap and lifeExp by continent/year

```{r summary}
gapminder_summary <- gapminder %>%
  group_by(continent, year) %>%
  summarise(mean_gdpPercap = mean(gdpPercap),
            mean_lifeExp = mean(lifeExp),
            .groups = "drop")

gapminder_summary
```

## 15: Line + point plot of year vs mean life expectancy

```{r plot3}
plot3 <- ggplot(gapminder_summary, aes(x = year, y = mean_lifeExp, colour = continent)) +
  geom_line() +
  geom_point()
plot3
```

## 16: kenya_gapminder

```{r kenya}
kenya_gapminder <- gapminder %>%
  filter(country == "Kenya")
```

## 17: Year vs GDP per capita for Kenya

```{r year v gdp}
ggplot(kenya_gapminder, aes(x = year, y = gdpPercap)) +
  geom_line(colour = "blue") +
  labs(x = "Year", y = "GDP per Capita",
       title = "Kenya: GDP per Capita Over Time") +
  theme_bw()
```

## 18: gapminder_decline

```{r decline}
gapminder_decline <- gapminder %>%
  filter(year %in% c(1997, 2002, 2007), diff_lifeExp < 0)
```

## 19: Bar plot of decline, faceted by year

```{r bar plot decline}
ggplot(gapminder_decline, aes(x = country, y = diff_lifeExp)) +
  geom_col() +
  facet_grid(~year, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
#The countries with the greatest reduction in life expectancy were:
# 1997 - Zimbabwe (-13.57 years)
# 2002 - Lesotho (-10.97 years)
# 2007 - Swaziland (-4.26 years)
```

## 20: Colour by country, remove legend, label axes, add theme, saved as plot4

```{r colour by country}
plot4 <- ggplot(gapminder_decline, aes(x = country, y = diff_lifeExp, fill = country)) +
  geom_col() +
  facet_grid(~year, scales = "free_x") +
  labs(x = "Country", y = "Change in Life Expectancy",
       title = "Countries with Declining Life Expectancy (1997, 2002, 2007)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none")
plot4
```

## 21: Combining plot1-4 with patchwork (2 on top, 2 on bottom)

```{r, fig.width=10, fig.height=8}
(plot1 | plot2) / (plot3 | plot4)
```
